### 2021-09-24 revision/ch1intro.tex
getting close.

\chapter{Sediment transport and landscape evolution}
\label{ch:Introduction}

@@

To address these complications, I develop in this thesis a handful of new bedload transport models using methods adopted from statistical physics.
To provide context and motivation, the thesis begins by reviewing earlier approaches to bedload transport which my work is most related to, rephrasing works as necessary during this review to indicate common themes, highlight the shortcomings to be addressed, and show continuity with my own developments which follow in subsequent chapters.

@@

A basic problem in sediment transport theory is to predict the downstream movement of individual particles.
At first glance, this seems an elementary problem in general physics, but numerous challenges appear.
Particles moving as bedload are driven downstream by a turbulent fluid flow, but relationships between the fluid flow and the applied forces are known only approximately \citep{Michaelides1997a,Schmeeckle2007}.
Downstream movement is resisted by frictional encounters (skidding, pivoting, colliding) between moving particles and the bed, but since the bed is a granular surface, its geometry is difficult to characterize \citep{Gordon1972}, and the role of bed interactions is challenging to describe \citep{Sekine1992, Nino1998}.

@@

An early motivation to understand individual particle motions was to predict the efficiency of sediment transport measurements \citep{Ettema2004}, and this motivation still drives a great deal of research into individual particle motions today \citep{Hassan2017,Pretzlav2021}.
A common technique to estimate sediment transport is to seed a stream with tracer stones and track their progress downstream \citep{Einstein1937, Takayama1965, Pretzlav2021}.
In principle, tracers provide a proxy for the population of grains in a stream, so one can estimate virtual velocities of tracers, then multiply by an estimate of the number of grains available for motion to calculate the overall sediment flux \citep{Wilcock1997a,Ferguson2002}.
In practice, challenges arise due to the distinct behavior of tracers over the local, intermediate, global, and geomorphic timescales. Apparently, tracer virtual velocities depend on the observation time, in a phenomenon which has been called ``advective slowdown" \citep{Ferguson2002,Haschenburger2011, Haschenburger2013, Pelosi2016}. This obscures the relationship between measured tracer virtual velocities and actual sediment fluxes, exposing a need for further research into how exactly individual particle motions depend on observation scale and what processes generate advective slowdown.

@@

Equation \ref{eq:einlangevin} is a kind of dynamical equation representing how the position of a sediment grain evolves through time \citep{Kubo1978}, similar to Newtonian mechanics \citep{Goldstein1997} but with random driving term (eq. \ref{eq:einrando}).
\begin{figure}[!htbp]
	\includegraphics[width=\linewidth,keepaspectratio]{./figures/ch1/einsteinConcept.pdf}
	\caption{Panel (a) indicates the representation of Einstein's model as an idealized Poisson pulse noise, as indicated in eq. \ref{eq:einrando}, while panel (b) shows the ``stairstep" trajectories of sediment particles moving downstream through cycles of steps (which are instantaneous) and rests (which have mean duration $1/k_E$). }
	\label{fig:einsteinfig}
\end{figure}

The governing equation of the distribution $P(x,t)$ to find the particle at $x$ can be calculated as an ensemble average of $\delta(x-x(t))$ over all possible realizations of the noise \citep{Risken1989,Moss1989}. Different methods exist to compute such averages \citep{Hanggi1978, Hanggi1984a, Balakrishnan1993, VanDenBroeck1983}, but whatever the approach, the governing equation for the distribution comes out as
\be  \big(\ell \px \pt + k_E \ell \px + \pt \big)P(x,t) = 0. \label{eq:einmaster}\ee
This equation can be solved by standard methods (series solutions or transform calculus) \citep{Arfken1985,Prudnikov1992a} to reproduce the original result of \citet{Einstein1937} for the probability distribution of position of a sediment particle:
\be P(x,t) = \Big[\delta(x) e^{-k_E t} + e^{-k_E t - x/\ell} \sqrt{\frac{k_E t}{\ell x}}\mathcal{I}_1\Big(2 \sqrt{\frac{k_E x t}{\ell}}\Big)\Big]\theta(x)\theta(t) \label{eq:eindist} \ee
Here, $\mathcal{I}_\nu$ is a modified Bessel function of order $\nu$. 
The probability distribution eq. \ref{eq:eindist} fully characterizes the downstream movement of an individual particle alternating randomly through steps and rests.

This distribution displays both advection and diffusion. Advection occurs because particles move downstream. Diffusion occurs because step lengths and instants when entrainment events occur vary from one particle to the next.

One can calculate all moments of the position by multiplying eq. \ref{eq:einmaster} by $x^n$ and integrating over space. This gives the mean position of the particle
\be \langle x (t)  \rangle = k_E \ell t. \ee
This equation indicates that in Einstein's model, sediment grains move with virtual velocity $V_\text{virt} = k_E \ell$ given by the product of the single-particle entrainment rate $k_E$ and mean step length $\ell$.

@@

The advent of high speed camera experiments produced new insight into the local and intermediate ranges of bedload motion \citep{Abbott1977,Francis1973,Drake1988} which \citet{Einstein1937} probably did not intend for his model to describe.
In the local range, particles move with a fluctuating velocity due to the variable drag of the turbulent flow \citep{Lajeunesse2010,Fathel2015} and changes in the particle's height within the flow profile \citep{VanRijn1984,Wiberg1985}.
In the intermediate range, particle-bed collisions impart additional variability to sediment velocities \citep{Gordon1972,Martin2013}.
Einstein's instantaneous movement assumption obscures the timescales over which these processes occur.

Studies by \citet{Lisle1998}, and \citet{Lajeunesse2017} generalized the Einstein theory to include intermediate timescales. They approximated particle velocities as constant (neglecting fluctuations), and assumed that the movement times are exponentially distributed random variables just like resting times, but this time characterized by a deposition rate $k_D$, whose reciprocal is the average period of time a particle spends in motion between entrainment and deposition.
The analogue of Einstein's model equation \ref{eq:einlangevin} with a finite movement velocity $V$ can be written
\be \dot{x} = V\eta(t), \label{eq:lislelangevin}\ee
%where the noise is now a ``dichotomous Markov noise" \citep{VanDenBroeck1990,Bena2006}, which is essentially a random switch or telegraph signal that alternates between ``on" ($\eta(t) = 1$, meaning the particle is moving) and ``off" ($\eta(t) = 0$, meaning the particle is resting) \citep{Cox1965,Horsthemke1984, Masoliver1991, Masoliver1996}.
This noise is displayed in fig. \ref{fig:lislefig} panel (a). Some particle trajectories produced by eq. \ref{eq:lislelangevin} are displayed in fig. \ref{fig:lislefig} panel (b).

The governing equation of the position probability distribution $P(x,t) = \langle \delta(x-\int_0^t V\eta(t')dt') \rangle$ becomes \citep{Balakrishnan1993}
\be \big(\pt^2 + V \px \pt + k_E V \px + k \pt \big) P(x,t) = 0,\label{eq:lislemaster}\ee
where $k_E$ and $k_D$ are the entrainment and deposition rates, $V$ is the particle velocity during the motion state, and $k = k_E+k_D$. This partial differential equation is called an asymmetric telegrapher's equation \citep{Rossetto2018}, and although the symmetric analogue of this equation is well-studied \citep{Weiss2002a, Masoliver2017}, the asymmetric problem is not often encountered.
\begin{figure}[!htbp]
	\includegraphics[width=\linewidth,keepaspectratio]{./figures/ch1/lisleConcept.pdf}
	\caption{Panel (a) indicates the generalization of Einstein's model to include the interval of sediment motion between entrainment and deposition, now represented with dichotomous noise eq. \ref{eq:lislelangevin}, while panel (b) shows the tilted stair-step trajectories of sediment particles moving downstream in cycles of motion at velocity $V$ (with mean duration $1/k_D$) and rest (with mean duration $1/k_E$). }
	\label{fig:lislefig}
\end{figure}

For the initial condition that particles have a probability $k_E/k$ to start in motion, reflecting steady state conditions as far as alternation between motion and rest are concerned \citep[e.g.][]{Ancey2006}, the solution of eq. \ref{eq:lislemaster} is \citep{Lisle1998}
\begin{multline}  P(x,t) =  e^{-\chi-\tau} \Big[ \frac{k_E}{V}\delta(\tau) + \frac{k_E}{V} \sqrt{\frac{\chi}{\tau}}\mathcal{I}_1(2\sqrt{\chi\tau}) + \frac{k_D}{V}\mathcal{I}_0(2\sqrt{\chi\tau}) \\
	+ \frac{k_Ek_D}{kV}\sqrt{\frac{\tau}{\chi}}\mathcal{I}_1(2\sqrt{\chi\tau}) + \frac{k_Ek_D}{kV} \delta(\chi)
	\Big]\theta(\chi)\theta(\tau), \label{eq:lisledist}
\end{multline}
where $\chi = k_D x/V$ and $\tau = k_E(t-x/V)$ are shorthand notations.
Incorporating the duration of sediment motion, the mean position of the sediment grain remains linear in time: $\langle x (t) \rangle  = k_E V t/k$, representing movement with an effective velocity $V_\text{eff} = k_E V/k$, which is the fraction of time spent in motion ($k_E/k)$ \citep{Ancey2006} multiplied by the velocity during the motion state. In contrast, the variance develops a two range scaling.
Computing $\sigma_x^2$ provides
\be \sigma_x(t)^2 = \frac{2k_Ek_DV^2}{k^3}\Big( t + \frac{1}{k}e^{-k t} - \frac{1}{k}\Big) , \label{eq:lislevar}\ee
which is a non-trivial result. At short times, for $t\ll 1/k$, eq. \ref{eq:lislevar} shows diffusion $\sigma_x^2 \sim t^2$, which is a faster ``ballistic" rate of spreading than the Einstein model predicts \citep{Sokolov2012}. At long times ($t\gg 1/k$), the diffusion becomes normal again ($\sigma_x^2 = 2 D_\text{eff} t$), with an effective diffusion constant $ D_\text{eff} = k_E k_D V^2/k^3$. 
The rapid spreading of particles at short timescales occurs because some are completely stationary while others move with velocity $V$.

@@

Some authors have modeled local and intermediate ranges of individual particle trajectories by writing approximate Newtonian equations for the dynamics of individual particles and integrating them numerically. Early efforts applied time-averaged fluid forces linked to the logarithmic mean flow velocity profile \citep{Yalin1963,VanRijn1984}, and later efforts included collision models to modify particle velocities upon bed contact \citep{Wiberg1985,Sekine1992,Nino1998}.

Researchers eventually included realistic granular interactions in many-particle simulations of bedload transport using the discrete element method \citep{Cundall1979a, Haff1993}.
The early works utilizing this approach used a two dimensional domain with a highly simplified ``slab" flow model \citep{Haff1993}, while later works have included synthetic turbulence to drive particles \citep{Schmeeckle2003,Maurin2015} or clever reduced-complexity representations of the flow \citep{Clark2015,Clark2017}.

The state of the art within this category of sediment transport models is to simulate both fluid and particle phases simultaneously.
Particle-particle interactions are represented with the discrete element method, where particles are represented as spheres \citep{Cudnall1979} or polytopes \citep{Wachs2019} while forces are phrased as idealized spring-dashpot models.
The fluid flow is modeled either by large eddy simulation or direct numerical simulation of the Navier-Stokes equations.
A majority of these approaches simulate the fluid without including no-slip boundary conditions on the particles \citep[e.g.][]{Vowinckel2014,Schmeeckle2014,Gonzalez2017}, although works are beginning to include the fluid boundary conditions on individual particles with the immersed boundary method \citep[e.g.][]{Ji2013, Ji2014, Elghannay2017, Yousefi2020}.
These computational physics models produce impressive insight into the underlying granular and fluid mechanics producing bed load transport \citep{Frey2011}, but analytical models are nevertheless desired for further insight into the problem.

@@

Another category of models describe bedload particle velocities using Langevin equations \citep{Ancey2014,Fan2014}, whereby the particles dynamics are driven by Gaussian white noise \citep{Kubo1978}.

@@

\citet{Fan2014} set out to describe exponenential-distributed bedload particle velocities ($u$) with the Langevin equation
\be \dot{u}(t) = -\Delta \text{sgn}(u) + F + \sqrt{2D}\xi(t). \label{eq:fanlangevin}\ee
This equation drives the particle velocity by a fluid drag $F + \sqrt{2D} \xi(t)$, where $F$ is a constant, $D$ is a diffusivity that characterizes the magnitude of drag fluctuations, and $\xi(t)$ is a Gaussian white noise with unit variance and vanishing mean \citep{Gardiner1983}.
This drag is resisted by a heuristic particle friction term $-\Delta \text{sgn}(u)$, introduced as a proxy for particle-bed collisions.
The equation governing the probability distribution $P(u,t)$ of the particle velocity is called a Fokker-Planck equation, and this can be derived from eq. \ref{eq:fanlangevin} as \citep{Risken1989,VanKampen2007} 
\be \frac{\partial}{\partial t} P(u,t) = -\Delta\frac{\partial}{\partial u}\Big[ \text{sgn}(u) P \Big] + D \frac{\partial^2 P}{\partial u^2},\ee
implying that the steady-state velocity distribution ($\pt P(x,t) = 0$) provided by eq. \ref{eq:fanlangevin} is
\be P(u) = \frac{\Delta^2-F^2}{2\Delta D}\exp\Big(-\frac{-\Delta |u| + F u}{D}\Big).\ee
This is the (two-sided) exponential distribution observed in one subset of experiments.

\citet{Ancey2014} formulated a Langevin equation to describe the Gaussian velocity distributions observed in the other subset of experiments. They wrote for the streamwise velocity 
\be t_r \dot{u}(t) = -(U-u) + \sqrt{2D}\xi(t),\ee
where $U$ is the mean velocity of particles, $D$ characterizes the magnitude of velocity fluctuations, and $\xi(t)$ is again a Gaussian white noise with vanishing mean and unit variance. The timescale $t_r$ is a relaxation time over which velocity fluctuations decay. This time, the Fokker-Planck equation is
\be \frac{\partial}{\partial t} P(u,t) = -\frac{\partial}{\partial u}\Big[\frac{U-u}{t_r}P\Big] + \frac{D}{t_r^2}\frac{\partial^2 P}{\partial u^2},\ee 
and the steady state solution becomes
\be P(u) = \sqrt{\frac{t_r}{2\pi D}} \exp\Big(-\frac{t_r(u-U)^2}{2 D}\Big), \ee
which is the Gaussian velocity distribution from the other subset of the experiments.
These two models of bedload velocity distributions are summarized in fig. \ref{fig:fanAncey}.
\begin{figure}[!htbp]
	\includegraphics[width=\linewidth,keepaspectratio]{./figures/ch1/fanAncey.pdf}
	\caption{Panel (a) demonstrates the trajectory of a particle within the Fan et al model, where Coulomb friction impedes turbulent drag. Panel (b) shows the resulting exponential particle velocity distribution, having relatively large fluctuations with mode at $u=0$.
	Panel (c) shows the analogous trajectory from the Ancey et al model, where turbulent Stokes drag generates the Gaussian velocity distribution seen in (d). The latter has relatively narrow and symmetric fluctuations, with a positive mode well above $u=0$.}
	\label{fig:fanAncey}
\end{figure}
Although these descriptions describe two end-member sediment velocity distributions, Gaussian and exponential, the underlying mechanical reason why bedload transport has shown multiple velocity distributions remains an open question, and other distributions have been reported besides exponential and Gaussian \citep[e.g.][]{Houssais2012,Liu2019}.
A comprehensive model of the full range of bedload transport observations remains a desirable target.

\subsection{The definition of the flux}
% for some reason the pdf of this figure does not properly render 
\begin{figure}[!htbp]
	\includegraphics[width=\linewidth,keepaspectratio]{./figures/ch1/fluxDefinitions.png}
	\caption{Particle fluxes can be defined with either a control volume $V$ or a control surface $S$. Here, particles are shown moving from time $t-\Delta t$ (faint white particles) to time $t$ (grey particles). Each particle's downstream displacement $\Delta x_i$ is indicated. These displacements define the downstream particle velocities $u_i \approx \Delta x_i/\Delta t.$ At the instant $t$, the surface definition of the flux eq. \ref{eq:surfflux} gives $q = \frac{1}{L}(u_6+u_7).$, while the volume definition eq. \ref{eq:controlflux} gives $\Phi = \frac{1}{L}(u_1+\dots+u_5)$,
	showing the non-equivalence of these two definitions.}
	\label{fig:fluxdefs}
\end{figure}
Perhaps a surprising summary of bedload transport research is that no one definition of the sediment flux has even been agreed upon despite over a century of research \citep{Ballio2018}.
Today, there are two main complementary definitions of the bedload flux being applied in stochastic modeling. Other definitions are discussed comprehensively in \citet{Ballio2014,Ballio2018}.

The first definition is reminiscent of continuum mechanics and formulates the flux as a kind of current of sediment across a control surface $S$ \citep{Furbish2012,Heyman2016,Ballio2014}: 
\be q = \int_S c(\textbf{x},t)\textbf{u}\cdot d\textbf{S}. \label{eq:surfflux} \ee
This definition involves the volumetric concentration $c$ of particles and their velocites at the instant they cross the control surface. These quantities require careful interpretation because bedload particles do not constitute a continuous field \citep{Furbish2012a}. Typically one represents the concentration as a sum of indicator or Dirac delta functions \citep[e.g.][]{}.

@@

One uniquely successful strategy to relate Einstein's model to flow and sediment characteristics was developed by \citet{Charru2004,Charru2006}. This ``erosion-deposition model" phrases sediment transport as a mass balance within a control volume using Einstein's entrainment rate $E$ and the complementary deposition rate $D$. These model parameters are related to flow and sediment properties using scaling relations obtained from experiments \citep{Charru2004, Charru2006, Lajeunesse2010,Seizilles2014,Lajeunesse2015}. This approach can be characterized as a mixture of the Einstein and Bagnold strategies for its reliance on both probabilistic rates and scaling relations.

@@

To obtain realistically-wide fluctuations in particle activity, they introduced a positive feedback called ``collective entrainment", whereby the entrainment rate of grains increases in proportion to the number of moving grains in the volume \citep{Ancey2008,Heyman2013}.

>>>>

\chapter{Sediment transport and channel evolution}
\label{ch:Introduction}

@@

To address these complications, I develop in this thesis a handful of new bedload transport models using methods adopted from statistical physics.
To provide context and motivation, the thesis begins by reviewing earlier approaches to bedload transport which my work is most related to, rephrasing works as necessary to indicate common themes, highlight the shortcomings to be addressed, and show continuity with my own developments which follow in subsequent chapters.

@@

A basic problem in sediment transport modeling is to predict the downstream movement of individual particles.
At first glance, this seems an elementary problem in general physics, but numerous challenges appear.
Particles moving as bedload are driven downstream by a turbulent fluid flow, but relationships between the fluid flow and the applied forces are known only approximately \citep{Michaelides1997,Schmeeckle2007}.
Downstream movement is resisted by frictional encounters between moving particles and the bed (skidding, pivoting, colliding), but since the bed is a granular surface, its geometry is difficult to characterize \citep{Gordon1972}, and the role of bed interactions is challenging to describe \citep{Sekine1992, Nino1998}.

@@

An early motivation to understand individual particle motions was to predict the efficiency of sediment transport measurements \citep{Ettema2004}, and this motivation still drives a great deal of research into individual particle motions today \citep{Hassan2017,Pretzlav2021}.
A common technique to estimate sediment transport is to seed a stream with tracer stones and track their progress downstream \citep{Einstein1937, Takayama1965, Pretzlav2021}.
In principle, tracers provide a proxy for the population of grains in a stream, so one can estimate virtual velocities of tracers, then multiply by an estimate of the number of grains available for motion to calculate the overall sediment flux \citep{Wilcock1997a,Ferguson2002}.
In practice, challenges arise due to the distinct behavior of tracers over the local, intermediate, global, and geomorphic timescales. Apparently, tracer virtual velocities depend on the observation time, in a phenomenon which has been called ``advective slowdown" \citep{Ferguson2002,Haschenburger2011, Haschenburger2013, Pelosi2016}. This observation time dependence obscures the relationship between measured tracer virtual velocities and actual sediment fluxes, exposing a need for further research into how exactly individual particle motions depend on observation scale and what processes generate advective slowdown.

@@

Equation \ref{eq:einlangevin} is a dynamical equation representing how the position of a sediment grain evolves through time \citep{Kubo1978}, similar to Newtonian mechanics \citep{Goldstein1997} but with random driving term (Eq. \ref{eq:einrando}).
\begin{figure}[!htbp]
	\includegraphics[width=\linewidth,keepaspectratio]{./figures/ch1/einsteinConcept.pdf}
	\caption{Panel (a) indicates the representation of Einstein's model as an idealized Poisson pulse noise, as indicated in Eq. \ref{eq:einrando}, while panel (b) shows the ``stair-step" trajectories of sediment particles moving downstream through cycles of steps (which are instantaneous) and rests (which have mean duration $1/k_E$). }
	\label{fig:einsteinfig}
\end{figure}

The governing equation of the distribution $P(x,t)$ to find the particle at $x$ can be calculated as an ensemble average of $\delta(x-x(t))$ over all possible realizations of the noise \citep{Risken1989,Moss1989}. Different methods exist to compute such averages \citep{Hanggi1978, Hanggi1984a, Balakrishnan1993, VanDenBroeck1983}, but whatever the approach, the governing equation for the distribution comes out as
\be  \big(\ell \px \pt + k_E \ell \px + \pt \big)P(x,t) = 0. \label{eq:einmaster}\ee
This equation can be solved by standard methods (series solutions or transform calculus) \citep{Arfken1985,Prudnikov1992a} to reproduce the original result of \citet{Einstein1937} for the probability distribution of position of a sediment particle:
\be P(x,t) = \Big[\delta(x) e^{-k_E t} + e^{-k_E t - x/\ell} \sqrt{\frac{k_E t}{\ell x}}\mathcal{I}_1\Big(2 \sqrt{\frac{k_E x t}{\ell}}\Big)\Big]\theta(x)\theta(t) \label{eq:eindist} \ee
Here, $\mathcal{I}_\nu$ is a modified Bessel function of order $\nu$. 
The probability distribution Eq. \ref{eq:eindist} fully characterizes the downstream movement of an individual particle alternating randomly through steps and rests.

This distribution displays both advection and diffusion. Advection occurs because particles move downstream. Diffusion occurs because step lengths and entrainment times vary from one particle to the next.

One can calculate all moments of the position by multiplying Eq. \ref{eq:einmaster} by $x^n$ and integrating over space. This gives the mean position of the particle
\be \langle x (t)  \rangle = k_E \ell t. \ee
This equation indicates that in Einstein's model, sediment grains move with virtual velocity $V_\text{virt} = k_E \ell$ given by the product of the single-particle entrainment rate $k_E$ and mean step length $\ell$.

@@

The advent of high speed camera experiments produced new insight into the local and intermediate ranges of bedload motion \citep{Abbott1977,Francis1973,Drake1988} which \citet{Einstein1937} probably did not intend for his model to describe.
In the local range, particles move with a fluctuating velocity due to the variable drag of the turbulent flow \citep{Lajeunesse2010,Fathel2015} and changes in the particle height within the flow profile \citep{VanRijn1984,Wiberg1985}.
In the intermediate range, particle-bed collisions impart additional variability to sediment velocities \citep{Gordon1972,Martin2013}.
Einstein's instantaneous movement assumption obscures the timescales over which these processes occur.

Studies by \citet{Lisle1998}, and \citet{Lajeunesse2017} generalized the Einstein theory to include intermediate timescales. They approximated particle velocities as constant (neglecting fluctuations), and they assumed that the movement times are exponentially distributed random variables just like resting times, but this time characterized by a deposition rate $k_D$, whose reciprocal is the average period of time a particle spends in motion between entrainment and deposition.
The analogue of Einstein's model equation \ref{eq:einlangevin} with a finite movement velocity $V$ can be written
\be \dot{x} = V\eta(t), \label{eq:lislelangevin}\ee
where the noise is now a ``dichotomous Markov noise" \citep{VanDenBroeck1990,Bena2006}, which is essentially a random switch or telegraph signal that alternates between ``on" ($\eta(t) = 1$, meaning the particle is moving) and ``off" ($\eta(t) = 0$, meaning the particle is resting) \citep{Cox1965,Horsthemke1984, Masoliver1991, Masoliver1996}.
This noise is displayed in Fig. \ref{fig:lislefig} (a). Some particle trajectories produced by Eq. \ref{eq:lislelangevin} are displayed in Fig. \ref{fig:lislefig} (b).

The governing equation of the position probability distribution $P(x,t) = \langle \delta(x-\int_0^t V\eta(t')dt') \rangle$ becomes \citep[cf.][]{Balakrishnan1993}
\be \big(\pt^2 + V \px \pt + k_E V \px + k \pt \big) P(x,t) = 0,\label{eq:lislemaster}\ee
where $k_E$ and $k_D$ are the entrainment and deposition rates, $V$ is the particle velocity during the motion state, and $k = k_E+k_D$. This partial differential equation is called an asymmetric telegrapher's equation \citep{Rossetto2018}, and although the symmetric analogue of this equation is well-studied \citep{Weiss2002a, Masoliver2017}, the asymmetric problem is not often encountered.
\begin{figure}[!htbp]
	\includegraphics[width=\linewidth,keepaspectratio]{./figures/ch1/lisleConcept.pdf}
	\caption{Panel (a) indicates the generalization of Einstein's model to include the interval of sediment motion between entrainment and deposition, now represented with dichotomous noise Eq. \ref{eq:lislelangevin}, while panel (b) shows the tilted stair-step trajectories of sediment particles moving downstream in cycles of motion at velocity $V$ (with mean duration $1/k_D$) and rest (with mean duration $1/k_E$). }
	\label{fig:lislefig}
\end{figure}

For the initial condition that particles have a probability $k_E/k$ to start in motion, reflecting steady state conditions as far as alternation between motion and rest are concerned \citep[e.g.][]{Ancey2006}, the solution of Eq. \ref{eq:lislemaster} is \citep{Lisle1998}
\begin{multline}  P(x,t) =  e^{-\chi-\tau} \Big[ \frac{k_E}{V}\delta(\tau) + \frac{k_E}{V} \sqrt{\frac{\chi}{\tau}}\mathcal{I}_1(2\sqrt{\chi\tau}) + \frac{k_D}{V}\mathcal{I}_0(2\sqrt{\chi\tau}) \\
	+ \frac{k_Ek_D}{kV}\sqrt{\frac{\tau}{\chi}}\mathcal{I}_1(2\sqrt{\chi\tau}) + \frac{k_Ek_D}{kV} \delta(\chi)
	\Big]\theta(\chi)\theta(\tau), \label{eq:lisledist}
\end{multline}
where $\chi = k_D x/V$ and $\tau = k_E(t-x/V)$ are shorthand notations.
Incorporating the duration of sediment motion, the mean position of the sediment grain remains linear in time: $\langle x (t) \rangle  = k_E V t/k$, representing movement with a virtual velocity $V_\text{virt} = k_E V/k$, which is the fraction of time spent in motion ($k_E/k)$ \citep{Ancey2006} multiplied by the velocity during the motion state. In contrast, the variance develops a two range scaling.
Computing $\sigma_x^2$ provides
\be \sigma_x(t)^2 = \frac{2k_Ek_DV^2}{k^3}\Big( t + \frac{1}{k}e^{-k t} - \frac{1}{k}\Big) , \label{eq:lislevar}\ee
which is a non-trivial result. At short times, for $t\ll 1/k$, Eq. \ref{eq:lislevar} shows diffusion $\sigma_x^2 \sim t^2$, which is a faster ``ballistic" rate of spreading than the Einstein model predicts \citep{Sokolov2012}. At long times ($t\gg 1/k$), the diffusion becomes normal again ($\sigma_x^2 = 2 D_\text{eff} t$), with an effective diffusion constant $ D_\text{eff} = k_E k_D V^2/k^3$. 
The rapid spreading of particles at short timescales occurs because some are completely stationary while others move with velocity $V$.

@@

Some authors have modeled local and intermediate ranges of individual particle trajectories by writing approximate Newtonian equations for the dynamics of individual particles and integrating them numerically. Early efforts applied time-averaged fluid forces linked to a logarithmic mean flow velocity profile \citep{Yalin1963,VanRijn1984}, and later efforts included collision models to modify particle velocities upon bed contact \citep{Wiberg1985,Sekine1992,Nino1998}.

Researchers eventually included realistic granular interactions in many-particle simulations of bedload transport using the discrete element method \citep[e.g.][]{Cundall1979, Wachs2019}.
The early works utilizing this approach used a two dimensional domain with a simplified ``slab" flow geometry \citep{Haff1993,Gotoh1997}, while later works included synthetic turbulence to drive particles \citep{Schmeeckle2003,Maurin2015} or clever reduced-complexity representations of the flow \citep{Clark2015,Clark2017}.

The state of the art within this category of sediment transport models is to simulate both fluid and particle phases simultaneously.
Particle-particle interactions are represented with the discrete element method, while the fluid flow is modeled either by large eddy simulation or direct numerical simulation of the Navier-Stokes equations.
A majority of these approaches simulate the fluid without including the no-slip boundary conditions on individual particles \citep[e.g.][]{Vowinckel2014,Schmeeckle2014,Gonzalez2017}, although some works have resolved these particle-scale fluid boundary conditions with the immersed boundary method \citep[e.g.][]{Ji2013, Ji2014, Elghannay2017, Yousefi2020}.
These computational physics models produce impressive insight into the underlying granular and fluid mechanics producing bed load transport \citep{Frey2011}, but analytical models are nevertheless desired for further insight into the problem.

@@

Another category of models describes bedload particle velocities using Langevin equations \citep{Ancey2014,Fan2014}, whereby the particles dynamics are driven by Gaussian white noise \citep{Kubo1978}.

@@

\citet{Fan2014} set out to describe exponenentially-distributed bedload particle velocities with the Langevin equation
\be \dot{u}(t) = -\Delta \text{sgn}(u) + F + \sqrt{2D}\xi(t). \label{eq:fanlangevin}\ee
This equation drives the particle velocity $u$ by a fluid drag $F + \sqrt{2D} \xi(t)$, where $F$ is a constant, $D$ is a diffusivity that characterizes the magnitude of drag fluctuations, and $\xi(t)$ is a Gaussian white noise with unit variance and vanishing mean \citep{Gardiner1983}.
This drag is resisted by a heuristic Coulomb friction term $-\Delta \text{sgn}(u)$, introduced as a proxy for particle-bed collisions. This model may have been inspired by models of Brownian motion with Coulomb friction in the physics literature \citep[e.g.][]{DeGennes2005,Touchette2010,Menzel2011}.
The equation governing the probability distribution $P(u,t)$ of the particle velocity is called a Fokker-Planck equation, and this can be derived from Eq. \ref{eq:fanlangevin} as \citep{Risken1989,VanKampen2007} 
\be \frac{\partial}{\partial t} P(u,t) = -\Delta\frac{\partial}{\partial u}\Big[ \text{sgn}(u) P \Big] + D \frac{\partial^2 P}{\partial u^2},\ee
implying that the steady-state velocity distribution ($\pt P(x,t) = 0$) provided by Eq. \ref{eq:fanlangevin} is
\be P(u) = \frac{\Delta^2-F^2}{2\Delta D}\exp\Big(-\frac{-\Delta |u| + F u}{D}\Big).\ee
This is the (two-sided) exponential distribution observed in one subset of experiments.

\citet{Ancey2014} formulated a Langevin equation to describe the Gaussian velocity distributions observed in the other subset of experiments. They wrote for the streamwise velocity 
\be t_r \dot{u}(t) = -(U-u) + \sqrt{2D}\xi(t),\ee
where $U$ represents the mean velocity of particles, $D$ characterizes the magnitude of velocity fluctuations, and $\xi(t)$ is again a Gaussian white noise with vanishing mean and unit variance. The timescale $t_r$ is a relaxation time over which velocity fluctuations decay. This time, the Fokker-Planck equation is
\be \frac{\partial}{\partial t} P(u,t) = -\frac{\partial}{\partial u}\Big[\frac{U-u}{t_r}P\Big] + \frac{D}{t_r^2}\frac{\partial^2 P}{\partial u^2},\ee 
and the steady state solution becomes
\be P(u) = \sqrt{\frac{t_r}{2\pi D}} \exp\Big(-\frac{t_r(u-U)^2}{2 D}\Big), \ee
which is the Gaussian velocity distribution from the other subset of the experiments.
These two models of bedload velocity distributions are summarized in Fig. \ref{fig:fanAncey}.
\begin{figure}[!htbp]
	\includegraphics[width=\linewidth,keepaspectratio]{./figures/ch1/fanAncey.pdf}
	\caption{Panel (a) demonstrates the trajectory of a particle within the Fan et al model, where Coulomb friction impedes turbulent drag. Panel (b) shows the resulting exponential particle velocity distribution, having relatively large fluctuations with mode at $u=0$.
	Panel (c) shows the analogous trajectory from the Ancey et al model, where turbulent Stokes drag generates the Gaussian velocity distribution seen in (d). The latter has relatively narrow and symmetric fluctuations, with a positive mode well above $u=0$.}
	\label{fig:fanAncey}
\end{figure}
Although these descriptions describe two end-member sediment velocity distributions, Gaussian and exponential, the underlying mechanical reason why bedload transport has shown multiple velocity distributions remains an open question, and several experiments have revealed other distributions besides exponential and Gaussian \citep[e.g.][]{Houssais2012,Liu2019}.
A comprehensive model of the full range of bedload transport observations remains a desirable target.

\subsection{The definition of the flux}
% for some reason the pdf of this figure does not properly render 
\begin{figure}[!htbp]
	\includegraphics[width=\linewidth,keepaspectratio]{./figures/ch1/fluxDefinitions.png}
	\caption{Particle fluxes can be defined with a control volume $V$ or a control surface $S$. Here, particles are shown moving from time $t-\Delta t$ (faint white particles) to time $t$ (grey particles). Each particle's downstream displacement $\Delta x_i$ is indicated. These displacements define the downstream velocities $u_i \approx \Delta x_i/\Delta t.$ At the instant $t$, the surface definition of the flux Eq. \ref{eq:surfflux} gives $q = \frac{1}{L}(u_6+u_7).$, while the volume definition Eq. \ref{eq:controlflux} gives $\Phi = \frac{1}{L}(u_1+\dots+u_5)$,
	showing the non-equivalence of these two definitions.}
	\label{fig:fluxdefs}
\end{figure}
Perhaps a surprising summary of bedload transport research is that no one definition of the sediment flux has been agreed upon despite over a century of research \citep{Ballio2018}.
Today, there are two main complementary definitions of the bedload flux being applied in stochastic modeling (but see \citet{Ballio2014,Ballio2018} for other definitions).

The first definition is reminiscent of continuum mechanics and formulates the flux as a current of sediment across a control surface $S$ \citep{Furbish2012,Heyman2016,Ballio2014}: 
\be q = \int_S c(\textbf{x},t)\textbf{u}\cdot d\textbf{S}. \label{eq:surfflux} \ee
This definition involves the volumetric concentration $c$ of particles and their velocites at the instant they cross the control surface. These quantities require careful interpretation because bedload particles do not constitute a continuous field \citep{Furbish2012a}. Typically one represents the concentration as a sum of indicator or Dirac delta functions \citep[cf.][Sec. 8.5.1]{Gardiner1983}.

@@

One uniquely successful strategy to relate Einstein's model to flow and sediment characteristics was developed by \citet{Charru2004,Charru2006}. This ``erosion-deposition model" phrases sediment transport as a mass balance within a control volume using Einstein's entrainment rate $E$ and the complementary deposition rate $D$. These rates are related to flow and sediment properties using scaling relations obtained from experiments \citep{Charru2004, Charru2006, Lajeunesse2010,Seizilles2014,Lajeunesse2015}. This approach can be characterized as a mixture of the Einstein and Bagnold strategies for its incorporate of both probabilistic rates and scaling relations.

@@

To obtain realistically-wide fluctuations in particle activity, Ancey et al. introduced a positive feedback they called ``collective entrainment", whereby the entrainment rate of grains increases in proportion to the number of moving grains in the volume \citep{Ancey2008,Heyman2013}.
